# Reproducible Research: Peer Assessment 1

## Loading and preprocessing the data
### Loading
Unzip and load `activity.csv`.
```{r load}
unzip("activity.zip")
activity <- read.csv("activity.csv")
```
### Preprocessing
Initially, the date column is loaded as Factors. 
We need to convert the date column to dates.
```{r castDate}
activity$date <- as.Date(activity$date, format = "%Y-%m-%d")
```

## What is mean total number of steps taken per day?
### Histogram of total number of steps taken each day
```{r hist}
totalSteps <- aggregate(steps ~ date, data=activity, sum, na.rm=TRUE)
hist(totalSteps$steps, col="red", xlab="Total Steps Per Day", ylab="Frequency",  main="Histogram of Total Steps Taken Per Day", breaks=10)
```

### Mean and median of total number of steps taken per day
```{r meanAndMedian}
meanTotSteps <- mean(totalSteps$steps, na.rm=TRUE)
medianTotSteps <- median(totalSteps$steps, na.rm=TRUE)
```

The mean total number of steps is `r meanTotSteps` and the median total number of steps is `r medianTotSteps`.


## What is the average daily activity pattern?
### Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r timeseries}
intervalSteps <- aggregate(steps ~ interval, activity, mean, na.rm=TRUE)
plot(intervalSteps$interval, intervalSteps$steps, type='l', col=1, 
     main="Average Number of Steps Taken, Averaged Over All Days", xlab="Interval", 
     ylab="Average number of steps")
```

### The 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps
```{r intervalWithMaxSteps}
maxIntervalStepsIndex <- which.max(intervalSteps$steps)
intervalWithMaxSteps <- intervalSteps[maxIntervalStepsIndex, "interval"]
maxSteps <- intervalSteps[maxIntervalStepsIndex, "steps"]
```

Interval `r intervalWithMaxSteps` contains the maximum number of steps, `r maxSteps`.

## Imputing missing values
### Total number of missing values in the dataset
```{r missingValues}
missingValuesCount <- sum(is.na(activity))
```

There are `r missingValuesCount` missing values in the dataset.

### Use median steps per interval to fill in missing values

```{r imputation}
# create dataframe with median steps per interval
medianStepsPerInterval <- aggregate(activity$steps, by=list(activity$interval), median, na.rm=TRUE)
names(medianStepsPerInterval)[1] <- "interval"
names(medianStepsPerInterval)[2] <- "medianSteps"

# merge activity and medianStepsPerInterval
missing <- merge(activity, medianStepsPerInterval, by="interval", sort=FALSE) 

# sort on date and interval
missing <- missing[with(missing, order(date, interval)), ]  

# replace NA in steps column with value in medianSteps column
missing$steps[is.na(missing$steps)] <- missing$medianSteps[is.na(missing$steps)]

# remove medianSteps column 
missing$medianSteps <- NULL

# missing values are now filled with median
newActivity <- missing[, c(2, 3, 1)]
head(newActivity)
```

### Histogram of the total number of steps taken each day from imputed steps using median
```{r imputedValuesHistogram}
totalImpSteps <- aggregate(newActivity$steps, by=list(newActivity$date), sum)
names(totalImpSteps)[1] <- "day"
names(totalImpSteps)[2] <- "steps"

hist(totalImpSteps$steps, col="red", main="Histogram of Total Steps Per Day (Imputed Steps)", 
    xlab="Total Steps Per Day", breaks=10)
```

### Mean and median of total number of steps taken per day from imputed steps using median
```{r meanAndMedianOfImputedSteps}
meanTotImpSteps <- mean(totalImpSteps$steps, na.rm=TRUE)
medianTotImpSteps <- median(totalImpSteps$steps, na.rm=TRUE)
```

The mean total number of steps is `r meanTotImpSteps` and the median total number of steps is `r medianTotImpSteps`.

Mean and median has changed after imputation. The mean has increased by `r meanTotSteps - meanTotImpSteps` and the median has increased by `r medianTotSteps - medianTotImpSteps`. The overall shape of the histogram has not changed except for a dramatic increase in low steps bin. 


## Are there differences in activity patterns between weekdays and weekends?
### Create a new factor variable in the dataset with two levels – “weekday” and “weekend”
```{r newFactorWeekdayAndWeekend}
# create a factor with the names of the days for all dates
newActivity$weekdays <- factor(format(newActivity$date, "%A"))
# the day names 
levels(newActivity$weekdays)
# replace the levels
levels(newActivity$weekdays) <- list(weekday=c("Monday", "Tuesday", "Wednesday", 
    "Thursday", "Friday"), weekend=c("Saturday", "Sunday"))
```

### Time series plot of the 5-minute interval vs. the average number of steps taken, averaged across all weekday days or weekend days.

```{r weekendAndWeekdayTimeSeries}
stepsMeanInterval <- aggregate(newActivity$steps, by=list(newActivity$weekdays, newActivity$interval), mean, na.rm=TRUE, na.action=NULL)
names(stepsMeanInterval)[1] <- "weekday"
names(stepsMeanInterval)[2] <- "interval"
names(stepsMeanInterval)[3] <- "meanSteps"

library(lattice)
xyplot(stepsMeanInterval$meanSteps ~ stepsMeanInterval$interval | 
    stepsMeanInterval$weekday, layout = c(1, 2), type = "l", xlab = "Interval", 
    ylab = "Number of steps")
```

